---
title: 'Text analytics HW: Wikipedia'
author: "Anna Prades"
date: "4 juliol de 2015"
output: html_document
---
In this assignment we will attempt to develop a vandalism detector that uses machine learning to distinguish between a valid edit and vandalism.

The data for this problem is based on the revision history of the page Language. Wikipedia provides a history for each page that consists of the state of the page at each revision. Rather than manually considering each revision, a script was run that checked whether edits stayed or were reverted. If a change was eventually reverted then that revision is marked as vandalism. This may result in some misclassifications, but the script performs well enough for our needs.

As a result of this preprocessing, some common processing tasks have already been done, including lower-casing and punctuation removal. The columns in the dataset are:

   * Vandal = 1 if this edit was vandalism, 0 if not.
   * Minor = 1 if the user marked this edit as a "minor edit", 0 if not.
   * Loggedin = 1 if the user made this edit while using a Wikipedia account, 0 if they did not.
   * Added = The unique words added.
   * Removed = The unique words removed.

Notice the repeated use of unique. The data we have available is not the traditional bag of words - rather it is the set of words that were removed or added. For example, if a word was removed multiple times in a revision it will only appear one time in the "Removed" column.

Load the data wiki.csv with the option stringsAsFactors=FALSE, calling the data frame "wiki". Convert the "Vandal" column to a factor using the command wiki$Vandal = as.factor(wiki$Vandal).

How many cases of vandalism were detected in the history of this page?

```{r}
wiki <- read.csv("wiki.csv", stringsAsFactors = FALSE)
str(wiki)
table(wiki$Vandal) #1815
#Codi proposed in the solution -> gives the same result, but converts 1st Vandal as factor (why?) 
wiki$Vandal = as.factor(wiki$Vandal)
table(wiki$Vandal) #1815
```

##Bag of words
We will now use the bag of words approach to build a model. We have two columns of textual data, with different meanings. For example, adding rude words has a different meaning to removing rude words. We'll start like we did in class by building a document term matrix from the Added column. **The text already is lowercase and stripped of punctuation.** So to pre-process the data, just complete the following four steps:

1) Create the corpus for the Added column, and call it "corpusAdded".

2) Remove the English-language stopwords.

3) Stem the words.

4) Build the DocumentTermMatrix, and call it dtmAdded.

If the code length(stopwords("english")) does not return 174 for you, then please run the line of code in this file, which will store the standard stop words in a variable called sw. When removing stop words, use tm_map(corpusAdded, removeWords, sw) instead of tm_map(corpusAdded, removeWords, stopwords("english")).

How many terms appear in dtmAdded?
```{r}
library(tm)
#Create a corpus
corpusAdded <- Corpus(VectorSource(wiki$Added))
corpusAdded = tm_map(corpusAdded, PlainTextDocument)
#Remove stopping words and stem the words
corpusAdded = tm_map(corpusAdded, removeWords, stopwords("english"))
strwrap(corpusAdded[[1]])

corpusAdded = tm_map(corpusAdded, stemDocument)
strwrap(corpusAdded[[1]])
#Create DocumentMatrix
dtmAdded = DocumentTermMatrix(corpusAdded)
dtmAdded

length(stopwords("english"))  #174

```

Filter out sparse terms by keeping only terms that appear in 0.3% or more of the revisions, and call the new matrix sparseAdded. How many terms appear in sparseAdded?
```{r}
sparseAdded = removeSparseTerms(dtmAdded, 0.997) #any term that it si not at least at 0.3% of the doc
sparseAdded
```

## Create a data frama called wordsAdded
Convert sparseAdded to a data frame called wordsAdded, and then prepend all the words with the letter A, by using the command:


```{r}
wordsAdded <- as.data.frame(as.matrix(sparseAdded))
colnames(wordsAdded) = paste("A", colnames(wordsAdded))
```

## Create a Removed bag-or words
Now repeat all of the steps we've done so far (create a corpus, remove stop words, stem the document, create a sparse document term matrix, and convert it to a data frame) to create a Removed bag-of-words dataframe, called wordsRemoved, except this time, prepend all of the words with the letter R:

colnames(wordsRemoved) = paste("R", colnames(wordsRemoved))

How many words are in the wordsRemoved data frame?

```{r}
#Create a corpus
corpusRemoved <- Corpus(VectorSource(wiki$Removed))

#Remove stopping words and stem the words
corpusRemoved <- tm_map(corpusRemoved, removeWords, stopwords("english"))
corpusRemoved <- tm_map(corpusRemoved, stemDocument)

#Create DocumentMatrix
dtmRemoved = DocumentTermMatrix(corpusRemoved)
dtmRemoved


# Remove sparse termes #any term that it si not at least at 0.3% of the doc
sparseRemoved = removeSparseTerms(dtmRemoved, 0.997)
sparseRemoved

#Create a data.frame, and past R 
wordsRemoved <- as.data.frame(as.matrix(sparseRemoved))
colnames(wordsRemoved) <- paste("R", colnames(wordsRemoved))
# How many words are in the wordsRemoved data frame?
str(wordsRemoved) #162 #or 
ncol(wordsRemoved)

```

##Combine the two data frames into a data frame called wikiWords
Then add the Vandal column (HINT: remember how we added the dependent variable back into our data frame in the Twitter lecture). Set the random seed to 123 and then split the data set using sample.split from the "caTools" package to put 70% in the training set.
What is the accuracy on the test set of a baseline method that always predicts "not vandalism" (the most frequent outcome)?
```{r}
wikiWords = cbind(wordsAdded, wordsRemoved) #there is a warning that som row names duplicated!!
# Add dependent variable
wikiWords$Vandal = wiki$Vandal
table(wikiWords$Vandal)

# Creating training and test sets
library(caTools)
set.seed(123)
spl <- sample.split(wikiWords$Vandal, SplitRatio = 0.7)
train <- subset(wikiWords, spl == TRUE)
test <- subset(wikiWords, spl == FALSE)

# Computing baseline accuracy on the test set
table(test$Vandal)
prop.table(table(test$Vandal)) #0.5313844
```

#Bag of words
Build a CART model to predict Vandal, using all of the other variables as independent variables. Use the training set to build the model and the default parameters (don't set values for minbucket or cp).

What is the accuracy of the model on the test set, using a threshold of 0.5? (Remember that if you add the argument type="class" when making predictions, the output of predict will automatically use a threshold of 0.5.)
```{r}
library(rpart)
library(rpart.plot)

vandalCART <- rpart(Vandal ~ ., data=train, method="class")
prp(vandalCART) #I do not understand this tree...

# Evaluate the performance of the model
predictCART = predict(vandalCART, newdata=test, type="class") #1st argument actual outcome, 2nd our predictions

t1 <- table(test$Vandal, predictCART)
t1
accuracy <- sum(diag(t1))/margin.table(t1)
accuracy #0.5417025. Just a little better thant the baseline


```

#Problem specific knowledge
We weren't able to improve on the baseline using the raw textual information. More specifically, the words themselves were not useful. There are other options though, and in this section we will try two techniques - identifying a key class of words, and counting words.

The key class of words we will use are website addresses. "Website addresses" (also known as URLs - Uniform Resource Locators) are comprised of two main parts. An example would be "http://www.google.com". The first part is the protocol, which is usually "http" (HyperText Transfer Protocol). The second part is the address of the site, e.g. "www.google.com". We have stripped all punctuation so links to websites appear in the data as one word, e.g. "httpwwwgooglecom". We hypothesize that given that a lot of vandalism seems to be adding links to promotional or irrelevant websites, the presence of a web address is a sign of vandalism.

We can search for the presence of a web address in the words added by searching for "http" in the Added column. The grepl function returns TRUE if a string is found in another string, e.g.

grepl("cat","dogs and cats",fixed=TRUE) # TRUE

grepl("cat","dogs and rats",fixed=TRUE) # FALSE

Create a copy of your dataframe from the previous question:

wikiWords2 = wikiWords

Make a new column in wikiWords2 that is 1 if "http" was in Added:

wikiWords2$HTTP = ifelse(grepl("http",wiki$Added,fixed=TRUE), 1, 0)

Based on this new column, how many revisions added a link?
```{r}
wikiWords2 <- wikiWords
wikiWords2$HTTP <- ifelse(grepl("http",wiki$Added,fixed=TRUE), 1, 0)
table(wikiWords2$HTTP)

```

## Problem specific knowledge
In problem 1.5, you computed a vector called "spl" that identified the observations to put in the training and testing sets. Use that variable (do not recompute it with sample.split) to make new training and testing sets:

wikiTrain2 = subset(wikiWords2, spl==TRUE)

wikiTest2 = subset(wikiWords2, spl==FALSE)

Then create a new CART model using this new variable as one of the independent variables.

What is the new accuracy of the CART model on the test set, using a threshold of 0.5?
```{r}
wikiTrain2 = subset(wikiWords2, spl==TRUE)
wikiTest2 = subset(wikiWords2, spl==FALSE)
#Create CART model with wikiWords2$HTTP
wikiCART2 <- rpart(Vandal ~ ., data = wikiTrain2, method = "class")
prp(wikiCART2)
#Prediction and accuracy in the test set
pred2 = predict(wikiCART2, newdata=wikiTest2, method="class")
pred.prob2 = pred2[,2]
t1 <- table(wikiTest2$Vandal, pred.prob2 >= 0.5)
t1
accuracy <- sum(diag(t1))/margin.table(t1)
accuracy #0.5726569
```

Another possibility is that the number of words added and removed is predictive, perhaps more so than the actual words themselves. We already have a word count available in the form of the document-term matrices (DTMs).

Sum the rows of dtmAdded and dtmRemoved and add them as new variables in your data frame wikiWords2 (called NumWordsAdded and NumWordsRemoved) by using the following commands:
```{r}
wikiWords2$NumWordsAdded = rowSums(as.matrix(dtmAdded))
wikiWords2$NumWordsRemoved = rowSums(as.matrix(dtmRemoved))
#What is the average number of words added?
summary(wikiWords2$NumWordsAdded)
```

In problem 1.5, you computed a vector called "spl" that identified the observations to put in the training and testing sets. Use that variable (do not recompute it with sample.split) to make new training and testing sets with wikiWords2. Create the CART model again (using the training set and the default parameters).

What is the new accuracy of the CART model on the test set?
```{r}
wikiTrain3 = subset(wikiWords2, spl==TRUE)
wikiTest3 = subset(wikiWords2, spl==FALSE)
#Create CART model with new variables( we have now http, and NumWordsAdded and Removed)
wikiCART3 <- rpart(Vandal ~ ., data = wikiTrain3, method = "class")
prp(wikiCART3)
#Prediction and accuracy in the test set
pred3 = predict(wikiCART3, newdata=wikiTest3, method="class")
pred.prob3 = pred3[,2]
t1 <- table(wikiTest3$Vandal, pred.prob3 >= 0.5)
t1
accuracy <- sum(diag(t1))/margin.table(t1)
accuracy # 0.6552021 Now it has improved!!
```

## Problem 3.1. Using non textual data
We have two pieces of "metadata" (data about data) that we haven't yet used. Make a copy of wikiWords2, and call it wikiWords3:
```{r}
wikiWords3 = wikiWords2
```

Then add the two original variables Minor and Loggedin to this new data frame:
```{r}
wikiWords3$Minor = wiki$Minor

wikiWords3$Loggedin = wiki$Loggedin
```

In problem 1.5, you computed a vector called "spl" that identified the observations to put in the training and testing sets. Use that variable (do not recompute it with sample.split) to make new training and testing sets with wikiWords3.

Build a CART model using all the training data. What is the accuracy of the model on the test set?
```{r}
wikiTrain4 = subset(wikiWords3, spl==TRUE)
wikiTest4 = subset(wikiWords3, spl==FALSE)
#Create CART model with new variables( we have now http, and NumWordsAdded and Removed)
wikiCART4 <- rpart(Vandal ~ ., data = wikiTrain4, method = "class")
prp(wikiCART4) #The chart, as the previous ones, it is not very enlightening... Only one split.
#Prediction and accuracy in the test set
pred4 = predict(wikiCART4, newdata=wikiTest4, method="class")
pred.prob4 = pred4[,2]
t1 <- table(wikiTest4$Vandal, pred.prob4 >= 0.5)
t1
accuracy <- sum(diag(t1))/margin.table(t1)
accuracy # 0.7188306
```

