---
title: "HW1 Document Clustering"
author: "Anna Prades"
date: "Friday, July 10, 2015"
output: html_document
---

document clustering with daily kos

Document clustering, or text clustering, is a very popular application of clustering algorithms. A web search engine, like Google, often returns thousands of results for a simple query. For example, if you type the search term "jaguar" into Google, around 200 million results are returned. This makes it very difficult to browse or find relevant information, especially if the search term has multiple meanings. If we search for "jaguar", we might be looking for information about the animal, the car, or the Jacksonville Jaguars football team. 

Clustering methods can be used to automatically group search results into categories, making it easier to find relavent results. This method is used in the search engines PolyMeta and Helioid, as well as on FirstGov.gov, the official Web portal for the U.S. government. The two most common algorithms used for document clustering are Hierarchical and k-means. 

In this problem, we'll be clustering articles published on Daily Kos, an American political blog that publishes news and opinion articles written from a progressive point of view. Daily Kos was founded by Markos Moulitsas in 2002, and as of September 2014, the site had an average weekday traffic of hundreds of thousands of visits. 

The file dailykos.csv contains data on 3,430 news articles or blogs that have been posted on Daily Kos. These articles were posted in 2004, leading up to the United States Presidential Election. The leading candidates were incumbent President George W. Bush (republican) and John Kerry (democratic). Foreign policy was a dominant topic of the election, specifically, the 2003 invasion of Iraq. 

Each of the variables in the dataset is a word that has appeared in at least 50 different articles (1,545 words in total). The set of  words has been trimmed according to some of the techniques covered in the previous week on text analytics (punctuation has been removed, and stop words have been removed). For each document, the variable values are the number of times that word appeared in the document. 

#Problem 1.1 HIERARCHICAL CLUSTERING
Let's start by building a hierarchical clustering model. First, read the data set into R. Then, compute the distances (using method="euclidean"), and use hclust to build the model (using method="ward.D"). You should cluster on all of the variables.

Running the dist function will probably take you a while. Why? Select all that apply.

```{r}
daily <- read.csv("dailykos.csv")
head(daily)
str(daily)
#Compute the distances
distance = dist(daily, method = "euclidean") #it takes a lot of time because it has a lot of variables and observations

#build the hclust model
clusterDaily = hclust(distance, method="ward")
# Plot the dendrogram
plot(clusterDaily)
#plot rectangles in order to figure up which number is better -> change k=n
rect.hclust(clusterDaily, k = 7, border = "red") #choices 2 or 3 ok, 5 or 6 too small distance
```

While looking at the dendogram 2 or 3 clusters seem a better choice..

**Thinking about the application**, it is probably better to show the reader more categories than 2 or 3. These categories would probably be too broad to be useful. Seven or eight categories seems more reasonable.

Let's pick 7 clusters. This number is reasonable according to the dendrogram, and also seems reasonable for the application. Use the cutree function to split your data into 7 clusters.

Now, we don't really want to run tapply on every single variable when we have over 1,000 different variables. Let's instead use the subset function to subset our data by cluster. Create 7 new datasets, each containing the observations from one of the clusters.

How many observations are in cluster 3?

```{r}
clusterGroups = cutree(clusterDaily, k = 7)
#Split the data by cluster
hierCluster <- split(daily, clusterGroups)
#Caulculate how many of them are there
sapply(hierCluster, nrow)
#Alternatively
cluster1 <- subset(daily, clusterGroups ==1)
nrow(cluster1)
cluster2 <- subset(daily, clusterGroups ==2)
nrow(cluster2)
cluster3 <- subset(daily, clusterGroups ==3)
nrow(cluster3)
cluster4 <- subset(daily, clusterGroups ==4)
nrow(cluster4)
cluster5 <- subset(daily, clusterGroups ==5)
nrow(cluster5)
cluster6 <- subset(daily, clusterGroups ==6)
nrow(cluster6)
cluster7 <- subset(daily, clusterGroups ==7)
nrow(cluster7)
```

Instead of looking at the average value in each variable individually, we'll just look at the top 6 words in each cluster. To do this for cluster 1, type the following in your R console :

```{r}
tail(sort(colMeans(cluster1)))
```
This computes the mean frequency values of each of the words in cluster 1, and then outputs the 6 words that occur the most frequently. The colMeans function computes the column (word) means, the sort function orders the words in increasing order of the mean values, and the tail function outputs the last 6 words listed, which are the ones with the largest column means.

Repeat this for the other clusters
```{r}
tail(sort(colMeans(cluster2)))
tail(sort(colMeans(cluster3)))
tail(sort(colMeans(cluster4)))
tail(sort(colMeans(cluster5)))
tail(sort(colMeans(cluster6)))
tail(sort(colMeans(cluster7)))
```

#Problem 2: k-means clustering
Now, run k-means clustering, setting the seed to 1000 right before you run the kmeans function. Again, pick the number of clusters equal to 7. You don't need to add the iters.max argument.

Subset your data into the 7 clusters (7 new datasets) by using the "cluster" variable of your kmeans output.

How many observations are in Cluster 3?

```{r}
set.seed(1000)
k=7
KMC = kmeans(daily, centers = k)
str(KMC)
#Subset the data into clusters and calculat the number of observations in each
kCluster <- split(daily, KMC$cluster)
sapply(kCluster, nrow)
#Alternatively, you could answer these questions by
table(KMC$cluster)
```
Now, output the six most frequent words in each cluster, like we did in the previous problem, for each of the k-means clusters.

```{r}
Kcluster1 <- subset(daily, KMC$cluster ==1)
Kcluster2 <- subset(daily, KMC$cluster ==2)
Kcluster3 <- subset(daily, KMC$cluster ==3)
Kcluster4 <- subset(daily, KMC$cluster ==4)
Kcluster5 <- subset(daily, KMC$cluster ==5)
Kcluster6 <- subset(daily, KMC$cluster ==6)
Kcluster7 <- subset(daily, KMC$cluster ==7)
#Output the 6 most frequent words
tail(sort(colMeans(Kcluster1)))
tail(sort(colMeans(Kcluster2)))
tail(sort(colMeans(Kcluster3)))
tail(sort(colMeans(Kcluster4)))
tail(sort(colMeans(Kcluster5)))
tail(sort(colMeans(Kcluster6)))
tail(sort(colMeans(Kcluster7)))

```

For the rest of this problem, we'll ask you to compare how observations were assigned to clusters in the two different methods. Use the table function to compare the cluster assignment of hierarchical clustering to the cluster assignment of k-means clustering.

Which Hierarchical Cluster best corresponds to K-Means Cluster 2? ->7
```{r}
table(clusterGroups, KMC$cluster)
round(prop.table(table(clusterGroups, KMC$cluster))*100, 3)
```

