---
title: "HW 3 Predictin Stock returns"
author: "Anna Prades"
date: "11 juliol de 2015"
output: html_document
---

In the second lecture sequence this week, we heard about cluster-then-predict, a methodology in which you first cluster observations and then build cluster-specific prediction models. In the lecture sequence, we saw how this methodology helped improve the prediction of heart attack risk. In this assignment, we'll use cluster-then-predict to predict future stock prices using historical stock data.

When selecting which stocks to invest in, investors seek to obtain good future returns. In this problem, we will first use clustering to identify clusters of stocks that have similar returns over time. Then, we'll use logistic regression to predict whether or not the stocks will have positive future returns.

For this problem, we'll use StocksCluster.csv, which contains monthly stock returns from the NASDAQ stock exchange. The NASDAQ is the second-largest stock exchange in the world, and it lists many technology companies. The stock price data used in this problem was obtained from infochimps, a website providing access to many datasets.

Each observation in the dataset is the monthly returns of a particular company in a particular year. The years included are 2000-2009. The companies are limited to tickers that were listed on the exchange for the entire period 2000-2009, and whose stock price never fell below $1. So, for example, one observation is for Yahoo in 2000, and another observation is for Yahoo in 2001. Our goal will be to predict whether or not the stock return in December will be positive, using the stock returns for the first 11 months of the year.

This dataset contains the following variables:

   * ReturnJan = the return for the company's stock during January (in the year of the observation). 
   * ReturnFeb = the return for the company's stock during February (in the year of the observation). 
   * ReturnMar = the return for the company's stock during March (in the year of the observation). 
   * ReturnApr = the return for the company's stock during April (in the year of the observation). 
  *  ReturnMay = the return for the company's stock during May (in the year of the observation). 
   * ReturnJune = the return for the company's stock during June (in the year of the observation). 
  *  ReturnJuly = the return for the company's stock during July (in the year of the observation). 
  *  ReturnAug = the return for the company's stock during August (in the year of the observation). 
  *  ReturnSep = the return for the company's stock during September (in the year of the observation). 
  *  ReturnOct = the return for the company's stock during October (in the year of the observation). 
  *  ReturnNov = the return for the company's stock during November (in the year of the observation). 
  *  PositiveDec = whether or not the company's stock had a positive return in December (in the year of the observation). This variable takes value 1 if the return was positive, and value 0 if the return was not positive.

For the first 11 variables, the value stored is a proportional change in stock value during that month. For instance, a value of 0.05 means the stock increased in value 5% during the month, while a value of -0.02 means the stock decreased in value 2% during the month.

#1. EXPLORING THE DATASET
Load StocksCluster.csv into a data frame called "stocks". How many observations are in the dataset?
```{r}
stocks <- read.csv("StocksCluster.csv")
dim(stocks)
```
What proportion of the observations have positive returns in December?
```{r}
table(stocks$PositiveDec)
prop.table(table(stocks$PositiveDec)) #Alternatively we could have used:mean(stocks$PositiveDec) 
```

What is the maximum correlation between any two return variables in the dataset? You should look at the pairwise correlations between ReturnJan, ReturnFeb, ReturnMar, ReturnApr, ReturnMay, ReturnJune, ReturnJuly, ReturnAug, ReturnSep, ReturnOct, and ReturnNov.
```{r}
cor(stocks)
round(cor(stocks), 3)
```
Which month (from January through November) has the largest mean return across all observations in the dataset? 7
```{r}
sapply(stocks, mean)
t1 <- sapply(stocks[ ,1:11], mean)
which.max(t1)
which.min(t1)
```

#2. LOGISTIC REGRESSION MODEL
Run the following commands to split the data into a training set and testing set, putting 70% of the data in the training set and 30% of the data in the testing set:
```{r}
library(caTools)
set.seed(144)
spl = sample.split(stocks$PositiveDec, SplitRatio = 0.7)
stocksTrain = subset(stocks, spl == TRUE)
stocksTest = subset(stocks, spl == FALSE)
# use the stocksTrain to train a logistic regression model (name it StocksModel) to predict PositiveDec using all the other variables as independent variables.
StocksModel <- glm(PositiveDec ~ ., data = stocksTrain, family ="binomial")
```
What is the overall **accuracy on the training set**, using a threshold of 0.5?
```{r}
predTrain <- predict(StocksModel, type="response")
t1 <- table(stocksTrain$PositiveDec, predTrain > 0.5)
t1
accuracy <- sum(diag(t1))/margin.table(t1)
accuracy #0.5711818
```
##Test set predictions
Now obtain test set predictions from StocksModel. What is the overall accuracy of the model on the test, again using a threshold of 0.5?
```{r}
#Test set baseline
prop.table(table(stocksTest$PositiveDec)) #0.5460564
#accuracy on the test set
predTest <- predict(StocksModel, newdata= stocksTest, type="response")
t1 <- table(stocksTest$PositiveDec, predTest > 0.5)
t1
accuracy <- sum(diag(t1))/margin.table(t1)
accuracy # 0.5670697
```

#PROBLEM 3. CLUSTERING STOCKS
Now, let's cluster the stocks. The first step in this process is to remove the dependent variable using the following commands:
```{r}
limitedTrain = stocksTrain
limitedTrain$PositiveDec = NULL

limitedTest = stocksTest
limitedTest$PositiveDec = NULL
```

Why do we need to remove the outcome variable?  In cluster-then-predict, our final goal is to predict the dependent variable, which is unknown to us at the time of prediction. Therefore, if we need to know the outcome value to perform the clustering, the methodology is no longer useful for prediction of an unknown outcome value.

This is an important point that is sometimes mistakenly overlooked. If you use the outcome value to cluster, you might conclude your method strongly outperforms a non-clustering alternative. However, this is because it is using the outcome to determine the clusters, which is not valid. 

##Normalizing the training set
(Why only the training set?)
In the market segmentation assignment in this week's homework, you were introduced to the preProcess command from the caret package, which normalizes variables by subtracting by the mean and dividing by the standard deviation.

In cases where we have a training and testing set, we'll want to normalize by the mean and standard deviation of the variables in the training set. We can do this by passing just the training set to the preProcess function:
```{r}
library(caret)
#The first command pre-processes the data, and the second command performs the normalization.
preproc = preProcess(limitedTrain)
normTrain = predict(preproc, limitedTrain)
normTest = predict(preproc, limitedTest)
#What is the mean of the ReturnJan variable in normTrain?
mean(normTrain$ReturnJan)
#What is the mean of the ReturnJan variable in normTest?
mean(normTest$ReturnJan)
```
Why is the mean ReturnJan variable much closer to 0 in normTrain than in normTest?
The distribution of the ReturnJan variable is different in the training and testing set. From mean(stocksTrain$ReturnJan) and mean(stocksTest$ReturnJan), we see that the average return in January is slightly higher in the training set than in the testing set. Since normTest was constructed by subtracting by the mean ReturnJan value from the training set, this explains why the mean value of ReturnJan is slightly negative in normTest. 
## k -means
Set the random seed to 144 (it is important to do this again, even though we did it earlier). Run k-means clustering with 3 clusters on normTrain, storing the result in an object called km.

Which cluster has the largest number of observations?
```{r}
set.seed(144)
k=3
km = kmeans(normTrain, centers = k, iter.max = 1000)
str(km)
table(km$cluster)
```
Recall from the recitation that we can use the flexclust package to obtain training set and testing set cluster assignments for our observations (note that the call to as.kcca may take a while to complete):
install.packages("flexclust")
```{r}
library(flexclust)
#We will not run the k-means algorith again -> we will apply the results on the test set
#flexclust has the KCCA, which stands for K-Centroids cluster analysis
# 1st argument original variable that stored information of the k-means
# 2nd input -> the data we clusterd
km.kcca = as.kcca(km, normTrain) 

clusterTrain = predict(km.kcca)
# now we can cluster the stocks in the normTest using the predict function
clusterTest = predict(km.kcca, newdata=normTest)
#How many test-set observations were assigned to Cluster 2?
table(clusterTest)
```

#PROBLEM 4. CLUSTER SPECIFIC PREDICTIONS
Using the subset function, build data frames stocksTrain1, stocksTrain2, and stocksTrain3, containing the elements in the stocksTrain data frame assigned to clusters 1, 2, and 3, respectively (be careful to take subsets of stocksTrain, not of normTrain). Similarly build stocksTest1, stocksTest2, and stocksTest3 from the stocksTest data frame.

Which training set data frame has the highest average value of the dependent variable?
```{r}
stocksTrain1 <- subset(stocksTrain, clusterTrain==1)
stocksTrain2 <- subset(stocksTrain, clusterTrain==2)
stocksTrain3 <- subset(stocksTrain, clusterTrain==3)
stocksTest1 <- subset(stocksTest, clusterTest==1)
stocksTest2 <- subset(stocksTest, clusterTest==2)
stocksTest3 <- subset(stocksTest, clusterTest==3)
mean(stocksTrain1$PositiveDec)
mean(stocksTrain2$PositiveDec)
mean(stocksTrain3$PositiveDec)

#Alternatively
trainCluster <- split(stocksTrain, clusterTrain)
sapply(trainCluster, colMeans)
testCluster <- split(stocksTest, clusterTest)
sapply(testCluster, colMeans)
```
##Build logistic regression model for each cluster in the training set
Build logistic regression models StocksModel1, StocksModel2, and StocksModel3, which predict PositiveDec using all the other variables as independent variables. StocksModel1 should be trained on stocksTrain1, StocksModel2 should be trained on stocksTrain2, and StocksModel3 should be trained on stocksTrain3.
```{r}
StocksModel1 <- glm(PositiveDec ~ ., data=stocksTrain1, family ="binomial")
StocksModel2 <- glm(PositiveDec ~ ., data=stocksTrain2, family ="binomial")
StocksModel3 <- glm(PositiveDec ~ ., data=stocksTrain3, family ="binomial")
summary(StocksModel1)
summary(StocksModel2)
summary(StocksModel3)
```
Which variables have a positive sign for the coefficient in at least one of StocksModel1, StocksModel2, and StocksModel3 and a negative sign for the coefficient in at least one of StocksModel1, StocksModel2, and StocksModel3? Select all that apply.
There are a lot of variables that change the sign of the coefficient. (Why?)
##Making test set predictions
Using StocksModel1, make test-set predictions called PredictTest1 on the data frame stocksTest1. Using StocksModel2, make test-set predictions called PredictTest2 on the data frame stocksTest2. Using StocksModel3, make test-set predictions called PredictTest3 on the data frame stocksTest3.
```{r}
PredictTest1 <- predict(StocksModel1, newdata=stocksTest1, type="response")
PredictTest2 <- predict(StocksModel2, newdata=stocksTest2, type="response")
PredictTest3 <- predict(StocksModel3, newdata=stocksTest3, type="response")

#Overall accuracy of StocksModel1 on the test set stocksTest1, using a threshold of 0.5
t1 <- table(stocksTest1$PositiveDec, PredictTest1 > 0.5)
t1
accuracy <- sum(diag(t1))/margin.table(t1)
accuracy # 0.6194145

#Overall accuracy of StocksModel2 on the test set stocksTest2, using a threshold of 0.5
t1 <- table(stocksTest2$PositiveDec, PredictTest2 > 0.5)
t1
accuracy <- sum(diag(t1))/margin.table(t1)
accuracy #0.5504808

#Overall accuracy of StocksModel3 on the test set stocksTest3, using a threshold of 0.5
t1 <- table(stocksTest3$PositiveDec, PredictTest3 > 0.5)
t1
accuracy <- sum(diag(t1))/margin.table(t1)
accuracy #0.6458333
```

## Computing the overall test set accuracy
To compute the overall test-set accuracy of the cluster-then-predict approach, we can combine all the test-set predictions into a single vector and all the true outcomes into a single vector:
```{r}
AllPredictions = c(PredictTest1, PredictTest2, PredictTest3)
AllOutcomes = c(stocksTest1$PositiveDec, stocksTest2$PositiveDec, stocksTest3$PositiveDec)

#Overall test set accuracy of the cluster-then-predict approach
t1 <- table(AllOutcomes, AllPredictions > 0.5)
t1
accuracy <- sum(diag(t1))/margin.table(t1)
accuracy 

```
We see a modest improvement over the original logistic regression model. Since predicting stock returns is a notoriously hard problem, this is a good increase in accuracy. By investing in stocks for which we are more confident that they will have positive returns (by selecting the ones with higher predicted probabilities), this cluster-then-predict model can give us an edge over the original logistic regression model.
